---
title: "edr4r: a live environmental data walkthrough"
subtitle: "Vector features, time series, rasters, trajectories, plots, and maps from three EDR services"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: false
    code-tools: true
    embed-resources: true
    smooth-scroll: true
    link-external-newwindow: true
execute:
  echo: true
  warning: false
  message: false
  fig-width: 9
  fig-height: 5
---

## What this notebook does

Every dataset in this document is requested from a live EDR endpoint when the notebook renders. There are no generated observations, saved fixtures, or local data snapshots. If an endpoint is unavailable, the render stops instead of substituting another dataset.

The walkthrough uses one small study area in the northern North Carolina Piedmont and three services:

1. USGS waterdata for streamgage vectors and recent daily discharge;
2. the Western Water Datahub for PRISM precipitation grids and a monthly temperature series;
3. Met Office Labs for a population grid and an elevation trajectory.

The common extent makes it possible to compare the returned data and place the station observations over both live raster sources.

::: {.callout-warning title="This notebook makes network requests"}
Public EDR services can change or become temporarily unavailable. Render the notebook shortly before a presentation. The request bounds, record limits, page limits, and batch limits below keep each pull explicit.
:::

```{r}
#| label: setup
#| include: false

options(width = 100, scipen = 6)

find_project_root <- function() {
  input <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) "")
  candidates <- unique(c(
    getwd(),
    dirname(getwd()),
    if (nzchar(input)) dirname(input) else character(),
    if (nzchar(input)) file.path(dirname(input), "..") else character()
  ))
  candidates <- normalizePath(candidates, mustWork = FALSE)
  hit <- candidates[file.exists(file.path(candidates, "DESCRIPTION"))]
  if (!length(hit)) stop("Could not find the edr4r project root.")
  hit[[1]]
}

project_root <- find_project_root()

# Use the package source in this checkout, which may be newer than CRAN.
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(project_root, export_all = FALSE, quiet = TRUE)
} else {
  library(edr4r)
}

library(ggplot2)
library(sf)

theme_set(theme_minimal(base_size = 12))

study_bbox <- c(-79.00, 35.70, -78.00, 36.30)
study_ring <- rbind(
  c(study_bbox[1], study_bbox[2]),
  c(study_bbox[3], study_bbox[2]),
  c(study_bbox[3], study_bbox[4]),
  c(study_bbox[1], study_bbox[4])
)

render_started <- Sys.time()
```

## 1. Connect to three live services

Creating a client does not contact the service. It records the base URL, timeout, retry policy, headers, and discovery-cache settings that every later request will use.

```{r}
#| label: clients

usgs <- edr_client(
  "https://api.waterdata.usgs.gov/ogcapi/beta",
  timeout = 60,
  max_tries = 2,
  cache_ttl = 300
)

wwdh <- edr_client(
  "https://api.wwdh.internetofwater.app",
  timeout = 60,
  max_tries = 2,
  cache_ttl = 300
)

met <- edr_client(
  "https://labs.metoffice.gov.uk/edr",
  timeout = 60,
  max_tries = 1,
  cache_ttl = 300
)

usgs
wwdh
met
```

## 2. Discover before retrieving data

The collection documents tell us which query types and parameters each service currently advertises. These are live metadata requests.

```{r}
#| label: live-discovery

usgs_collections <- edr_collections(usgs, refresh = TRUE)
prism_capabilities <- edr_capabilities(wwdh, "usgs-prism", refresh = TRUE)
population_capabilities <- edr_capabilities(
  met,
  "global_pop_density",
  refresh = TRUE
)
terrain_capabilities <- edr_capabilities(
  met,
  "copernicus_dem",
  refresh = TRUE
)

collection_summary <- data.frame(
  service = c("USGS waterdata", "Western Water Datahub", "Met Office Labs", "Met Office Labs"),
  collection = c("daily-edr", "usgs-prism", "global_pop_density", "copernicus_dem"),
  query_used = c("locations", "cube and position", "area", "trajectory"),
  stringsAsFactors = FALSE
)

knitr::kable(collection_summary)
```

The PRISM collection advertises three parameters. We will use precipitation for a cube and maximum temperature for a position time series.

```{r}
knitr::kable(
  prism_capabilities$parameters[, c("id", "name", "unit_symbol")],
  col.names = c("Parameter ID", "Name", "Unit")
)
```

The query metadata can be checked before any data request.

```{r}
query_checks <- data.frame(
  collection = c("daily-edr", "usgs-prism", "usgs-prism", "global_pop_density", "copernicus_dem"),
  query = c("locations", "cube", "position", "area", "trajectory"),
  advertised = c(
    edr_supports(usgs, "daily-edr", query = "locations"),
    edr_supports(prism_capabilities, query = "cube"),
    edr_supports(prism_capabilities, query = "position"),
    edr_supports(population_capabilities, query = "area"),
    edr_supports(terrain_capabilities, query = "trajectory")
  )
)
knitr::kable(query_checks)
```

## 3. Live vector features from USGS

`edr_locations()` requests a GeoJSON station index and promotes it to `sf`. Pagination is allowed only inside the page and feature caps supplied here.

```{r}
#| label: live-usgs-locations

all_gauges <- edr_locations(
  usgs,
  "daily-edr",
  bbox = study_bbox,
  limit = 100,
  paginate = TRUE,
  max_pages = 10,
  max_features = 500
)

stream_candidates <- all_gauges[
  grepl("Stream", all_gauges$site_type, ignore.case = TRUE) &
    is.finite(all_gauges$drainage_area),
]
selection_index <- unique(round(seq(
  1,
  nrow(stream_candidates),
  length.out = min(12, nrow(stream_candidates))
)))
gauges <- stream_candidates[selection_index, ]

gauge_columns <- intersect(
  c("id", "monitoring_location_name", "county_name", "drainage_area", "geometry"),
  names(gauges)
)
cat(nrow(all_gauges), "locations were returned across the expanded study area.\n")
cat(nrow(gauges), "stream sites were selected for bounded observation requests.\n")
gauges[, gauge_columns]
```

This is a regular `sf` object. No spatial conversion is needed before using `ggplot2` or another `sf`-aware package.

```{r}
#| label: live-vector-plot

ggplot(gauges) +
  geom_sf(data = all_gauges, color = "#b8c2cc", size = 1.4, alpha = 0.6) +
  geom_sf(aes(size = drainage_area), color = "#236192", alpha = 0.9) +
  geom_sf_text(
    aes(label = sub(", NC$", "", monitoring_location_name)),
    nudge_y = 0.008,
    size = 2.8,
    check_overlap = TRUE
  ) +
  scale_size_continuous(range = c(2.5, 7)) +
  coord_sf(expand = FALSE) +
  labs(
    title = "USGS monitoring locations returned by EDR",
    subtitle = "Blue sites are selected for time-series retrieval",
    size = "Drainage area",
    x = NULL,
    y = NULL
  )
```

## 4. Live station time series from USGS

USGS currently ignores `datetime` on individual `daily-edr` location requests and returns the latest `limit` observations. The batch helper makes one bounded request per station and records the outcome of each request.

```{r}
#| label: live-usgs-series

streamflow_batch <- edr_location_batch(
  usgs,
  "daily-edr",
  location_id = gauges$id,
  parameter_name = "00060",
  limit = 90,
  max_requests = nrow(gauges),
  on_error = "collect",
  progress = FALSE
)

streamflow_batch$requests
streamflow_batch$errors

streamflow <- streamflow_batch$data
successful_ids <- streamflow_batch$requests$location_id[
  streamflow_batch$requests$status == "success"
]
gauges_with_data <- gauges[gauges$id %in% successful_ids, ]
streamflow$station <- gauges$monitoring_location_name[
  match(streamflow$.location_id, gauges$id)
]

streamflow[1:10, c(".location_id", "station", "parameter", "unit", "datetime", "value")]
```

`edr_plot()` detects the temporal domain and returns a regular `ggplot` object.

```{r}
#| label: live-usgs-series-plot

edr_plot(
  streamflow,
  parameter = "00060",
  group = "station",
  facet = NULL,
  geom = "line"
) +
  labs(
    title = "Recent daily discharge from USGS",
    subtitle = paste(length(successful_ids), "successful live station requests"),
    x = NULL,
    y = "Discharge (ft3/s)",
    color = "Station"
  )
```

## 5. Live PRISM raster request from WWDH

The `cube` query returns five full years of monthly precipitation grids from the live `usgs-prism` collection. This is 60 raster slices over the expanded study area, while the explicit bounds keep the request suitable for an interactive demonstration.

```{r}
#| label: live-prism-cube

prism_response <- edr_cube(
  wwdh,
  "usgs-prism",
  bbox = study_bbox,
  datetime = "2016-01-01/2020-12-01",
  parameter_name = "ppt"
)

prism_grid <- covjson_to_tibble(prism_response)

prism_response
prism_grid[1:10, c("parameter", "unit", "datetime", "x", "y", "value")]
```

The returned rows can be summarized without knowing the grid dimensions in advance. This table reports the actual temporal coverage and cells per monthly slice.

```{r}
prism_monthly <- aggregate(
  value ~ datetime + parameter + unit,
  data = prism_grid,
  FUN = function(x) mean(x, na.rm = TRUE)
)
names(prism_monthly)[names(prism_monthly) == "value"] <- "spatial_mean"

prism_summary <- data.frame(
  first_month = min(prism_grid$datetime),
  last_month = max(prism_grid$datetime),
  monthly_slices = length(unique(prism_grid$datetime)),
  cells_per_slice = nrow(prism_grid) / length(unique(prism_grid$datetime)),
  total_rows = nrow(prism_grid)
)
knitr::kable(prism_summary)
```

The spatial mean condenses each returned raster to one value, making the five-year time pattern easy to inspect while the full cell-level series remains available for mapping and analysis.

```{r}
#| label: live-prism-raster-series

ggplot(prism_monthly, aes(datetime, spatial_mean)) +
  geom_line(color = "#236192", linewidth = 0.75) +
  geom_point(color = "#236192", size = 1.4) +
  labs(
    title = "Five years of monthly PRISM raster slices",
    subtitle = "Each point is the spatial mean of one live raster response slice",
    x = NULL,
    y = "Study-area mean precipitation (mm/month)"
  )
```

### Static raster visualization

Selecting one returned month gives a conventional raster view.

```{r}
#| label: live-prism-static

prism_first_month <- prism_grid[
  prism_grid$datetime == min(prism_grid$datetime),
]

edr_plot(prism_first_month, view = "grid") +
  labs(
    title = "PRISM mean monthly precipitation",
    subtitle = format(min(prism_grid$datetime), "%B %Y", tz = "UTC"),
    x = "Longitude",
    y = "Latitude"
  )
```

### `terra` interoperability

The live tidy grid can be converted to a `SpatRaster` for raster algebra.

```{r}
#| label: live-prism-terra

if (requireNamespace("terra", quietly = TRUE)) {
  prism_raster <- terra::rast(
    prism_first_month[, c("x", "y", "value")],
    type = "xyz",
    crs = "EPSG:4326"
  )
  prism_raster
  terra::plot(
    prism_raster,
    col = hcl.colors(30, "Blues 3", rev = TRUE),
    main = "Live PRISM cells as a terra SpatRaster"
  )
} else {
  cat("Install terra to run this interoperability step.\n")
}
```

## 6. Live PRISM point time series

The same collection supports a `position` query. This request retrieves twelve monthly maximum-temperature observations at the center of the study area.

```{r}
#| label: live-prism-position

study_center <- c(
  mean(study_bbox[c(1, 3)]),
  mean(study_bbox[c(2, 4)])
)

prism_point_response <- edr_position(
  wwdh,
  "usgs-prism",
  coords = study_center,
  datetime = "2020-01-01/2020-12-01",
  parameter_name = "tmx"
)

prism_temperature <- covjson_to_tibble(prism_point_response)
prism_temperature[, c("parameter", "unit", "datetime", "x", "y", "value")]
```

```{r}
#| label: live-prism-position-plot

edr_plot(
  prism_temperature,
  group = NULL,
  facet = NULL,
  geom = "both"
) +
  labs(
    title = "Monthly maximum temperature from PRISM",
    subtitle = sprintf("Position %.2f, %.2f", study_center[1], study_center[2]),
    x = NULL,
    y = "Maximum temperature (degC)"
  )
```

## 7. Live population raster from Met Office Labs

The Met Office Labs `global_pop_density` collection supports an `area` query. `edr_area()` closes the coordinate matrix into a WKT polygon and sends it with an explicit response CRS.

```{r}
#| label: live-met-population

population_response <- edr_area(
  met,
  "global_pop_density",
  coords = study_ring,
  parameter_name = "Pop_Density",
  crs = "EPSG:4326"
)

population_grid <- covjson_to_tibble(population_response)

population_response
population_grid[1:10, c("parameter", "unit", "x", "y", "value")]
```

```{r}
#| label: live-met-population-plot

edr_plot(population_grid, view = "grid") +
  labs(
    title = "Population density from Met Office Labs",
    subtitle = paste(sum(is.finite(population_grid$value)), "returned cells with values"),
    x = "Longitude",
    y = "Latitude"
  )
```

## 8. Live elevation trajectory from Met Office Labs

The path is built from three live USGS station coordinates. The `copernicus_dem` trajectory query samples elevation along that path and returns a CoverageJSON `MultiPoint` domain.

```{r}
#| label: live-met-trajectory

gauge_xy <- st_coordinates(gauges_with_data)
gauge_xy <- gauge_xy[order(gauge_xy[, "X"]), , drop = FALSE]
path_index <- unique(round(seq(1, nrow(gauge_xy), length.out = 3)))
path_coords <- gauge_xy[path_index, c("X", "Y"), drop = FALSE]

terrain_response <- edr_trajectory(
  met,
  "copernicus_dem",
  coords = path_coords,
  parameter_name = "Height",
  crs = "EPSG:4326"
)

terrain_path <- covjson_to_tibble(terrain_response)

terrain_response
terrain_path[c(1:5, (nrow(terrain_path) - 4):nrow(terrain_path)),
             c("parameter", "unit", "x", "y", "value")]
```

Distance along the path is derived from the returned coordinates after projection to a metric CRS.

```{r}
#| label: live-met-trajectory-plot

terrain_sf <- st_as_sf(
  terrain_path,
  coords = c("x", "y"),
  crs = 4326,
  remove = FALSE
)
terrain_xy_m <- st_coordinates(st_transform(terrain_sf, 5070))
terrain_path$distance_km <- c(
  0,
  cumsum(sqrt(rowSums(diff(terrain_xy_m)^2))) / 1000
)

ggplot(terrain_path, aes(distance_km, value)) +
  geom_area(fill = "#8c6d31", alpha = 0.35) +
  geom_line(color = "#5c3b16", linewidth = 0.7) +
  labs(
    title = "Copernicus elevation along a live EDR trajectory",
    x = "Distance along path (km)",
    y = "Elevation above mean sea level (m)"
  )
```

## 9. Interactive maps

### USGS stations with time-series and CSV popups

Click a station to inspect its returned daily discharge or download its rows as CSV.

```{r}
#| label: live-station-map

station_map <- edr_map(
  gauges,
  data = streamflow,
  popup = "all",
  id_col = "id",
  label_col = "monitoring_location_name",
  location_col = ".location_id",
  parameter = "00060"
)
station_map
```

### PRISM grid with a month selector

Every returned month stays in this widget. Use the datetime control to switch slices and click a cell for its values.

```{r}
#| label: live-prism-map

prism_map <- edr_map(
  prism_grid,
  initial = list(
    parameter = "ppt",
    datetime = "2016-01-01 00:00:00"
  ),
  grid_transform = "sqrt",
  grid_opacity = 0.82
)
prism_map
```

### PRISM grid and USGS stations together

The grid and station series come from different live services. `edr_add_stations()` keeps the grid cells below the monitoring markers and gives the station layer its own control.

```{r}
#| label: live-composite-map

composite_map <- edr_map(
  prism_grid,
  initial = list(
    parameter = "ppt",
    datetime = "2016-01-01 00:00:00"
  ),
  grid_transform = "sqrt"
)

composite_map <- edr_add_stations(
  composite_map,
  locations = gauges,
  data = streamflow,
  popup = "plot+csv",
  id_col = "id",
  label_col = "monitoring_location_name",
  location_col = ".location_id",
  parameter = "00060",
  group = "USGS streamgages"
)

composite_map
```

### Met Office population grid and USGS stations

This second cross-service map uses the live Met Office population coverage as its raster layer.

```{r}
#| label: live-population-station-map

population_station_map <- edr_map(
  population_grid,
  parameter = "Pop_Density",
  grid_transform = "sqrt",
  grid_opacity = 0.78
)

population_station_map <- edr_add_stations(
  population_station_map,
  locations = gauges,
  data = streamflow,
  popup = "plot+csv",
  id_col = "id",
  label_col = "monitoring_location_name",
  location_col = ".location_id",
  parameter = "00060",
  group = "USGS streamgages"
)

population_station_map
```

Each widget can be written as a separate, self-contained HTML file.

```{r}
#| eval: false

edr_save_html(composite_map, "prism-and-streamgages.html")
edr_save_html(population_station_map, "population-and-streamgages.html")
```

## 10. What was retrieved

This table is calculated from the live responses in this render.

```{r}
#| label: live-provenance-summary

retrieval_summary <- data.frame(
  service = c(
    "USGS waterdata",
    "USGS waterdata",
    "Western Water Datahub",
    "Western Water Datahub",
    "Met Office Labs",
    "Met Office Labs"
  ),
  collection = c(
    "daily-edr", "daily-edr", "usgs-prism", "usgs-prism",
    "global_pop_density", "copernicus_dem"
  ),
  query = c("locations", "location batch", "cube", "position", "area", "trajectory"),
  rows = c(
    nrow(all_gauges), nrow(streamflow), nrow(prism_grid), nrow(prism_temperature),
    nrow(population_grid), nrow(terrain_path)
  ),
  parameter = c("station metadata", "00060", "ppt", "tmx", "Pop_Density", "Height"),
  stringsAsFactors = FALSE
)

knitr::kable(retrieval_summary)
cat("Render started:", format(render_started, tz = "UTC", usetz = TRUE), "\n")
```

## 11. The rest of the EDR query vocabulary

The package also supports items, radius, and corridor queries. Whether a call works depends on the collection metadata, so check `edr_capabilities()` or `edr_supports()` first.

```{r}
query_reference <- data.frame(
  function_name = c(
    "edr_locations() / edr_location()",
    "edr_items() / edr_item()",
    "edr_position()",
    "edr_area()",
    "edr_cube()",
    "edr_radius()",
    "edr_trajectory()",
    "edr_corridor()"
  ),
  selection = c(
    "station index or one station id",
    "feature index or one feature id",
    "point or multipoint",
    "polygon or multipolygon",
    "2D or 3D bounding box",
    "point plus radius",
    "path through space",
    "path plus corridor width and height"
  ),
  common_result = c(
    "GeoJSON index, then CoverageJSON or CSV observations",
    "GeoJSON",
    "CoverageJSON point or profile",
    "CoverageJSON subset",
    "CoverageJSON grid or collection",
    "CoverageJSON subset",
    "CoverageJSON trajectory",
    "CoverageJSON corridor"
  )
)

knitr::kable(query_reference, col.names = c("Function", "Selection", "Common result"))
```

## Suggested walk-through

A 35-minute presentation can follow this order:

1. Show the three clients and the advertised capability checks.
2. Follow the USGS location response from GeoJSON to `sf` and a station map.
3. Pull the bounded station batch and open one time-series popup.
4. Run the WWDH PRISM cube request, inspect its tidy cells, and move through the map's monthly slices.
5. Convert one returned PRISM slice to `terra`.
6. Compare the Met Office population raster with the USGS station layer.
7. Finish with the elevation trajectory and the retrieval summary.

The document demonstrates the complete path from live service discovery to retrieval, parsing, analysis, plotting, and interactive mapping without introducing local observation data.
